---
title: "Simulation of Critical Care Allocation Strategies During the COVID-19 Pandemic"
output:
  html_document:
    toc: yes
    toc_depth: '2'
    df_print: paged
  html_notebook:
    toc: yes
    toc_depth: 2
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '2'
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r libraries, include=FALSE, cache=FALSE}
library(tidyverse)
library(furrr)
library(pROC)
library(grid)
library(lubridate)
library(flextable)
future::plan(multiprocess)
library(radiant)
library(ggpubr)
library(svglite)
```

#Transform data into usable format
```{r}

nmhc <- read_csv("/data/COVID/data801/SOFA_COVID/res_icu_data2.csv") 
uc = read_csv("/data/COVID/data801/SOFA_COVID/uc_deidentified.csv") 

#Combine de-identified vectors of age, race, comorbidity score, SOFA, outcome from both medical systems
covid_sample = rbind(nmhc, uc)

#Elixhauser score transform into comorbidity category (primary analysis of 75th and 90th percentile for major and severe)
major = unname(quantile(covid_sample$wscore_ahrq, c(0.75), na.rm=TRUE))
severe = unname(quantile(covid_sample$wscore_ahrq, c(0.90), na.rm=TRUE))

covid_sample = covid_sample%>%
  mutate(
    chronic_disease_state = case_when(
      wscore_ahrq < major|is.na(wscore_ahrq) ~ "none",
      wscore_ahrq >= major & wscore_ahrq < severe ~ "major",
      wscore_ahrq >= severe ~ "severe"
    ),
    chronic_disease_state = factor(chronic_disease_state, levels = c("none", "major", "severe")),
    alive = 1-died)

```


#Set simulation parameters
```{r}
set.seed(121456)
sample_size <- nrow(covid_sample)

N_samples <- 10000
sample_list <- vector(mode = "list", length = N_samples)

for (n in seq(1:N_samples)) {
  sample_list[[n]] <- sample_n(covid_sample, sample_size, replace = T)
}

degree_scarcity <- 0.5
vents <- floor(sample_size*degree_scarcity)
```


#Helper functions
```{r}
lives_saved <- function(allocation){
  allocation %>%
    filter(get_vent == "critical care (survival)") %>%
    nrow()
}



allocation_performance <- function(alloc_list){
  lives <- map_int(alloc_list, lives_saved)
  
  return(tibble(lives_saved = lives))
}
```

#Allocation systems

## Sickest first
"Sickest first" allocation represents the default in american medicine under non-triage conditions. The sickest patients are prioritized because the less severely ill patients can wait their turn.
```{r}
sickest_first <- function(sim_pop, num_vents = vents){
  
  allocate <- sim_pop %>%
    arrange(-sofa_num) %>%
    mutate(get_vent = factor(case_when(
      row_number() <= num_vents & alive ==1 ~  "critical care (survival)",
      row_number() <= num_vents ~ "critical care (death)",
      TRUE ~ "comfort care only"), levels = c("critical care (death)",
                                            "critical care (survival)",
                                            "comfort care only"))
    )
  return(allocate)
}



sickest_list <- map(.x = sample_list, sickest_first)

sickest_first <- allocation_performance(sickest_list)
```


## NY allocation
The New York allocation scheme relies on a categorization of the SOFA score and a lottery within priority tiers. Chronic disease state is not incorporated
```{r}
ny_allocate <- function(sim_pop,  num_vents = vents){
  
  lottery <- runif(n = sim_pop %>% nrow())
  
  
  allocate <- sim_pop %>%
    cbind(lottery) %>%
    mutate(triage_cat = case_when(
      sofa_num < 8 ~ "Highest",
      sofa_num < 12 ~ "Intermediate",
      TRUE ~ "No critical care"
    ), priority_score = case_when(
      triage_cat == "Highest" ~ lottery,
      triage_cat == "Intermediate" ~ 1 + lottery,
      TRUE ~ 2 + lottery)
    ) %>%
    arrange(priority_score) %>%
    mutate(get_vent = factor(case_when(
      row_number() <= num_vents & alive ==1 ~  "critical care (survival)",
      row_number() <= num_vents ~ "critical care (death)",
      TRUE ~ "comfort care only"), levels = c("critical care (death)",
                                            "critical care (survival)",
                                            "comfort care only"))
    )
  
  return(allocate)
  
  
}


ny_list <- map(sample_list, ny_allocate)

new_york <- allocation_performance(ny_list)
```

## Maryland
This framework uses a combination of SOFA tiers and points for severe chronic disease, with age as a tiebreaker.
```{r}
maryland_rank <- function(sim_pop, num_vents = vents){
  lottery <- runif(n = sim_pop %>% nrow())
  
  allocate <- sim_pop %>%
    cbind(lottery) %>%
    mutate(score = case_when(sofa_num < 9 ~ 1,
                             sofa_num < 12 ~ 2,
                             sofa_num < 15 ~ 3,
                             TRUE ~ 4) +
             case_when(
               chronic_disease_state == "severe" ~ 3,
               TRUE ~ 0),
           age_cat_2 = case_when(
             age < 50 ~ 1,
             age < 70 ~ 2,
             age < 85 ~ 3,
             TRUE ~ 4)
    ) %>%
    arrange(score, age_cat_2, lottery) %>%
    mutate(get_vent = factor(case_when(
      row_number() <= num_vents & alive ==1 ~  "critical care (survival)",
      row_number() <= num_vents ~ "critical care (death)",
      TRUE ~ "comfort care only"), levels = c("critical care (death)", "critical care (survival)", "comfort care only"))
    )
  
  return(allocate)
}

maryland_list <- future_map(sample_list, maryland_rank)
maryland <- allocation_performance(maryland_list)
```


## Pennsylvania
This framework uses a combination of SOFA tiers and points for major and severe chronic disease, with age as a tiebreaker.
```{r}
penn_pitt_allocate <- function(sim_pop, num_vents = vents){
  
  lottery <- runif(n = sim_pop %>% nrow())
  
  allocate <- sim_pop %>%
    cbind(lottery) %>%
    mutate(score = case_when(sofa_num < 6 ~ 1,
                             sofa_num < 9 ~ 2,
                             sofa_num < 12 ~ 3,
                             TRUE ~ 4) +
             case_when(chronic_disease_state == "major" ~ 2,
                       chronic_disease_state == "severe" ~ 4,
                       TRUE ~ 0),
           age_cat_2 = case_when(
             age < 41 ~ 1,
             age < 61 ~ 2,
             age < 76 ~ 3,
             TRUE ~ 4)
    ) %>%
    arrange(score, age_cat_2, lottery) %>%
    mutate(get_vent = factor(case_when(
      row_number() <= num_vents & alive ==1 ~  "critical care (survival)",
      row_number() <= num_vents ~ "critical care (death)",
      TRUE ~ "comfort care only"), levels = c("critical care (death)", "critical care (survival)", "comfort care only"))
    )
  
  return(allocate)
}

penn_pitt_list <- map(sample_list, penn_pitt_allocate)

penn_pitt <- allocation_performance(penn_pitt_list)
```

## Lottery allocation
In lottery allocation, critical care are assigned randomly without regard to the patient's age or probability of survival
```{r}
lottery_allocate <- function(sim_pop, num_vents = vents){
  
  lottery <- runif(n = sim_pop %>% nrow())
  
  allocate <- sim_pop %>%
    cbind(lottery) %>%
    arrange(lottery) %>%
    mutate(get_vent = factor(case_when(
      row_number() <= num_vents & alive ==1 ~  "critical care (survival)",
      row_number() <= num_vents ~ "critical care (death)",
      TRUE ~ "comfort care only"), levels = c("critical care (death)",
                                            "critical care (survival)",
                                            "comfort care only"))
    )
  
  return(allocate)
}

lottery_list <- map(sample_list, lottery_allocate)
lottery <-allocation_performance(lottery_list)
```

## Youngest first
In youngest first allocation, the patients with the lowest age are assigned critical cares first. The severity of illness of the patients are ignored.
```{r}
youngest_allocate <- function(sim_pop, num_vents = vents){
  
  allocate <- sim_pop %>%
    arrange(age) %>%
    mutate(get_vent = factor(case_when(
      row_number() <= num_vents & alive ==1 ~  "critical care (survival)",
      row_number() <= num_vents ~ "critical care (death)",
      TRUE ~ "comfort care only"), levels = c("critical care (death)",
                                            "critical care (survival)",
                                            "comfort care only"))
    )
  
  return(allocate)
}

youngest_list <- map(sample_list, youngest_allocate)
youngest <- allocation_performance(youngest_list)
```

# Comparing allocation systems
## Table 2 - Combined dataframe
```{r, cache =FALSE}
combined_df <- lottery %>%
  mutate(system = "Lottery") %>%
  rbind(sickest_first %>% mutate(system = "Sickest first")) %>%
  rbind(youngest %>% mutate(system = "Youngest first")) %>%
  rbind(new_york %>% mutate(system = "New York"))%>%
  rbind(maryland %>% mutate(system = "Maryland")) %>%
  rbind(penn_pitt %>% mutate(system = "Pennsylvania")) %>%
  mutate(system = factor(system, levels = c("Sickest first",
                                            "Lottery",
                                            "Youngest first",
                                            "New York",
                                            "Maryland",
                                            "Pennsylvania")))


table2 = combined_df%>%
  group_by(system)%>%
  summarise(lives = mean(100*lives_saved/sample_size),
            lives5 = quantile(100*lives_saved/sample_size, probs = 0.025) %>% as.numeric(),
            lives95 = quantile(100*lives_saved/sample_size, probs = 0.975) %>% as.numeric())


```

## Figure 2 - Distribution of lives-saved by system
```{r, echo=FALSE}
lives_saved_plot <- combined_df %>%
  ggplot(aes(x= system, fill = system, y =  100*lives_saved/sample_size)) + # color = system,
  # geom_dotplot(binaxis = "y", stackdir='center', binwidth =  0.003) +
  geom_violin(trim=F, adjust=5) +
  geom_boxplot(width=0.1, outlier.shape = NA) +
  # stat_summary(fun.data="mean_sdl", mult=1, geom="crossbar", width=1 ) +
  theme_classic() +
  theme(axis.text = element_text(angle = 45, hjust = 1)) +theme(legend.position = "none")+
  labs(x = "", y = "Lives saved (%)")

lives_saved_plot

ggsave("lives_saved.svg", plot = lives_saved_plot)


```


## Lives Saved Analysis
```{r, echo=FALSE}


data_wide <- combined_df%>%
  select(lives_saved, system)%>%
  mutate(subject = rep(1:N_samples, times=6))

data_wide = spread(data_wide, system, lives_saved)


variance = c()
lowcon = c()
highcon = c()
conf = c()
systems = c("Sickest first", "Lottery", "Youngest first", "New York", "Maryland", "Pennsylvania")
for (i in 1:6) {
  lowcon[i]=mean(combined_df$lives_saved[which(combined_df$system==systems[i])]) - (1.96*sd(combined_df$lives_saved[which(combined_df$system==systems[i])]))
  highcon[i]=mean(combined_df$lives_saved[which(combined_df$system==systems[i])]) + (1.96*sd(combined_df$lives_saved[which(combined_df$system==systems[i])]))
  conf[i] = paste(round(lowcon[i]*100/sample_size, digits = 1), "-", round(highcon[i]*100/sample_size, digits=1), sep = "")
}

lives_saved=table2%>%
  mutate(confidence_interval = conf)

lives_saved = lives_saved%>%
  mutate(credible_interval = paste(round(lives5, digits=1), "-", round(lives95, digits=1), sep = ""))


ttest = function(system1, system2){
  numerator = abs(mean((combined_df$lives_saved[which(combined_df$system==system1)])- (combined_df$lives_saved[which(combined_df$system==system2)])))
  variance_1= sd(combined_df$lives_saved[which(combined_df$system==system1)])^2
  variance_2 = sd(combined_df$lives_saved[which(combined_df$system==system2)])^2
  covar = data_wide%>%
    select(matches(system1)|matches(system2))%>%
    mutate(mean1 = mean(.[[1]]))%>%
    mutate(mean2 = mean(.[[2]]))%>%
    mutate(diff = abs(.[[1]]-mean1)*abs(.[[2]]-mean2))%>%
    summarise(sum = sum(diff))%>%
    deframe()/N_samples
  z = numerator/(variance_1 + variance_2 - 2*covar)^0.5
  p.value = 2*pt(-abs(z), df=sample_size-1)
  p.value = ifelse(p.value<0.001, "<0.001", paste("=", signif(p.value, digits=1), sep = ""))
  
  return(p.value)
}

lives_comparisons = as.data.frame(matrix(nrow=6, ncol=7))%>%
  mutate(V1= systems)%>%
  rename("Sickest first" = V2)%>%
  rename("Lottery" = V3)%>%
  rename("Youngest first" = V4)%>%
  rename("New York" = V5)%>%
  rename("Maryland" = V6)%>%
  rename("Pennsylvania" = V7)


for (i in 1:6) {
  for (j in 1:6) {
    lives_comparisons[i,j+1] = ttest(systems[i], systems[j])
  }
}


write_csv(lives_comparisons, "lives_comparison.csv")


```
